---
title: "Inflammation cleaning and reporting - example with EASE data"
author: "Michelle Byrne"
date: "March 6, 2018"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# Steps for cleaning, tidying, and reporting on inflammatory markers.
## This example is for salivary SIgA, IL-1 beta, and CRP from the EASE Study (PI: mbyrne@uoregon.edu)

Load libraries and read in csv data

```{r}
library(rio)
library(tidyverse)
library(psych)
library(robustHD)
library(ggplot2)
library(DescTools)
library(corrplot)
library(Hmisc)
library(moments)
setwd("A:/Studies/EASE/output/processed_immunology_data")
easeimm <- read_csv("immunology_data.csv")
View(easeimm)
str(easeimm)

```

# Step 1: Decide how to deal with missing data 
```{r}
# Take out completely missing samples (if you want to impute instead, do this beforehand & report)
# Listwise deletion option (BAD!):
# easeimm <- na.omit(easeimm$pid)
#Another option: Recode NaN to missing values (NA): 
easeimm [easeimm == "NaN"] <- NA

```

# Step 2: Deal with out of range values
```{r}
# Out of range values (OOR). These are values that the assay was not sensitive enough to give a value for. They are usually indicated by OOR> (out of range too high, i.e., right censored) or OOR< (out of range too low, i.e., left censored)

# Our consensus is that rather than treat these values as missing data (and either imputing or listwise deleting), these should also be winsorized the same way other outliers are. For example, changed to a value of the maximum value in the dataset plus one increment (or minimum minus one). An increment is the level that the assay can detect to, e.g., 0.01 pg/ml. The reasoning for this is that rather than these values representing an error, they are still valuable data that the assay is saying it can either not detect (too low), or can't estimate because they are too high. For further reference, see doi: 10.1016/j.bbi.2013.08.010

# Winsorizing actual outliers will be a step later in the process. For now, we will identify OOR cells and replace them with a number that is either the max or min of the variable in this dataset, plus the increment

increment = 0.001
oor_right = "OOR>"
oor_left = "OOR<"
num_oor_right_siga <- length(na.omit(easeimm$siga [easeimm$siga == oor_right]))
num_oor_right_il1b <- length(na.omit(easeimm$il1b [easeimm$il1b == oor_right]))
num_oor_right_crp <- length(na.omit(easeimm$crp [easeimm$crp == oor_right]))
num_oor_left_siga <- length(na.omit(easeimm$siga [easeimm$siga == oor_left]))
num_oor_left_il1b <- length(na.omit(easeimm$il1b [easeimm$il1b == oor_left]))
num_oor_left_crp <- length(na.omit(easeimm$crp [easeimm$crp == oor_left]))

# Report these in your manuscript:
num_oor_right_siga
num_oor_right_il1b
num_oor_right_crp
num_oor_left_siga
num_oor_left_il1b
num_oor_left_crp

# Now replace (winsorize) the OOR samples

# First identify what the max/min are outside of the OOR values (and force them to be numeric, because if they had OOR in them they wouldn't be):
siga_num <- easeimm[!(easeimm$siga==oor_right),]
siga_num$siga <- as.numeric(siga_num$siga)
max_siga <- max(na.omit(siga_num$siga))
min_siga <- min(na.omit(siga_num$siga))

il1b_num <- easeimm[!(easeimm$il1b==oor_right),]
il1b_num$il1b <- as.numeric(il1b_num$il1b)
max_il1b <- max(na.omit(il1b_num$il1b))
min_il1b <- min(na.omit(il1b_num$il1b))

crp_num <- easeimm[!(easeimm$crp==oor_right),] 
crp_num$crp <- as.numeric(crp_num$crp)
max_crp <- max(na.omit(crp_num$crp))
min_crp <- min(na.omit(crp_num$crp))

# If the immune marker is OOR> then replace with max plus increment, OOR< then replace with min minus increment

easeimm$siga <- ifelse (easeimm$siga==oor_right,max_siga+increment,easeimm$siga)
easeimm$il1b <- ifelse (easeimm$il1b==oor_right,max_il1b+increment,easeimm$il1b)
easeimm$crp <- ifelse (easeimm$crp==oor_right,max_crp+increment,easeimm$crp)

easeimm$siga <- ifelse (easeimm$siga==oor_left,max_siga-increment,easeimm$siga)
easeimm$il1b <- ifelse (easeimm$il1b==oor_left,max_il1b-increment,easeimm$il1b)
easeimm$crp <- ifelse (easeimm$crp==oor_left,max_crp-increment,easeimm$crp)

# Then force everything to be numeric for further analyses

easeimm$siga <- as.numeric(easeimm$siga)
easeimm$il1b <- as.numeric(easeimm$il1b)
easeimm$crp <- as.numeric(easeimm$crp)
str(easeimm)

```

# Step 3 (for saliva only): Correct your data for flow rate
#### Flow rate = *(Volume of saliva)/(time sample took in sec)*
#### Corrected concentration = *(observed concentration) x (Flow rate)*
##### CRP is not flow rate dependent
##### SIgA *is* flow rate dependent
##### IL-1 beta does not appear to be flow rate dependent in individuals with IL-1 beta levels in the normal range

```{r}
vol_ml = 2
easeimm$fr <- vol_ml / easeimm$flowsec
easeimm$siga_frc <- easeimm$siga * easeimm$fr
```


# Step 4: Visualize your data to see normality and outiers

Histograms for immune markers to visualize normality
```{r}
hist(easeimm$siga_frc)
hist(easeimm$crp)
hist(easeimm$il1b)
```

Boxplots to visualize outliers
```{r}

ggplot(easeimm, aes(x = "", y = siga_frc)) +   
  geom_boxplot() +
  ylab("SIgA FRC (pg/ml)") +
  ggtitle("EASE SIgA") +
  geom_smooth(method='lm', color="black")

ggplot(easeimm, aes(x = "", y = il1b)) +   
  geom_boxplot() +
  ylab("IL1-B (pg/ml)") +
  ggtitle("EASE IL-1B") +
  geom_smooth(method='lm', color="black")

ggplot(easeimm, aes(x = "", y = crp)) +   
  geom_boxplot() +
  ylab("CRP (pg/ml)") +
  ggtitle("EASE CRP") +
  geom_smooth(method='lm', color="black")
```

# Step 5. Calculate and report normality statistics (skew and kurtosis)
Kurtosis and Skew should be -2/+2 (West, et al. 1995)

```{r}
siga_sk <- skewness(easeimm$siga_frc, na.rm=TRUE)
il1b_sk <- skewness(easeimm$il1b, na.rm=TRUE)
crp_sk <- skewness(easeimm$crp, na.rm=TRUE)
siga_ku <- kurtosis(easeimm$siga_frc, na.rm=TRUE)
il1b_ku <- kurtosis(easeimm$il1b, na.rm=TRUE)
crp_ku <- kurtosis(easeimm$crp, na.rm=TRUE)
```

# Step 6. If necessary, transform data
Report the type of transformation (e.g., ln, log10, etc.)
Visualize boxplot again after transformation

```{r}
# Transform 
# if skew or kurtosis is < -2 or > 2 then calculate the natural log of that variable

if (siga_sk > 2 | siga_sk < -2 | siga_ku > 2 | siga_ku < -2) {
easeimm$ln_siga <- as.numeric(lapply(easeimm$siga_frc, log))
} 

if (il1b_sk > 2 | il1b_sk < -2 | il1b_ku > 2 | il1b_ku < -2) {
easeimm$ln_il1b <- as.numeric(lapply(easeimm$il1b, log))
}

if (crp_sk > 2 | crp_sk < -2 | crp_ku > 2 | crp_ku < -2) {
easeimm$ln_crp <- as.numeric(lapply(easeimm$crp, log))
}


# Check everything post transformation for any transformed variables. 

ln_siga_sk <- skewness(easeimm$ln_siga, na.rm = TRUE)
siga_sk
ln_siga_ku <- kurtosis(easeimm$ln_siga, na.rm = TRUE)
siga_ku

ln_il1b_sk <- skewness(easeimm$ln_il1b, na.rm = TRUE)
il1b_sk
ln_il1b_ku <- kurtosis(easeimm$ln_il1b, na.rm = TRUE)
il1b_ku

ln_crp_sk <- skewness(easeimm$ln_crp, na.rm = TRUE)
crp_sk
ln_crp_ku <- kurtosis(easeimm$ln_crp, na.rm = TRUE)
crp_ku

hist(easeimm$ln_siga)
hist(easeimm$ln_il1b)
hist(easeimm$ln_crp)


ggplot(easeimm, aes(x = "", y = ln_siga)) +   
  geom_boxplot() +
  ylab("log (SIgA FRC)") +
  ggtitle("EASE SIgA (log transformed)") +
  geom_smooth(method='lm', color="black")

ggplot(easeimm, aes(x = "", y = ln_il1b)) +   
  geom_boxplot() +
  ylab("log (IL1-B)") +
  ggtitle("EASE IL1-B (log transformed)") +
  geom_smooth(method='lm', color="black")

ggplot(easeimm, aes(x = "", y = ln_crp)) +   
  geom_boxplot() +
  ylab("log (CRP)") +
  ggtitle("EASE CRP (log transformed)") +
  geom_smooth(method='lm', color="black")

# If transforming makes the skew and kurtosis worse, don't use the transformed data (i.e.,  make sure not to use transformed data from this point on). DO make a note that you MAY need to use non-parametric testing for this variable instead. BUT wait until you have fixed the outliers below and checked skew/kurtosis first.

# In this example, only IL1-B got better from transforming, so we leave CRP and SIgA alone.


```

# Step 7. Dealing with outliers (includes previously OOR values)

```{r}

siga_mean <- mean(easeimm$siga_frc, na.rm = TRUE)
siga_sd <- sd(easeimm$siga_frc, na.rm = TRUE)
siga_up_limit <- (siga_mean + 3*(siga_sd))
siga_up_limit
siga_lo_limit <- (siga_mean - 3*(siga_sd))

il1b_mean <- mean(easeimm$ln_il1b, na.rm = TRUE)
il1b_sd <- sd(easeimm$ln_il1b, na.rm = TRUE)
il1b_up_limit <- (il1b_mean + 3*(il1b_sd))
il1b_up_limit
il1b_lo_limit <- (il1b_mean - 3*(il1b_sd))

crp_mean <- mean(easeimm$crp, na.rm = TRUE)
crp_sd <- sd(easeimm$crp, na.rm = TRUE)
crp_up_limit <- (crp_mean + 3*(crp_sd))
crp_up_limit
crp_lo_limit <- (crp_mean - 3*(crp_sd))


# But for the low limit we have to make sure it isn't impossible, i.e., negative
siga_lo_limit <- ifelse(siga_lo_limit<0, 0, siga_lo_limit)
siga_lo_limit

il1b_lo_limit <- ifelse(il1b_lo_limit<0, 0, il1b_lo_limit)
il1b_lo_limit

crp_lo_limit <- ifelse(crp_lo_limit<0, 0, crp_lo_limit)
crp_lo_limit

# in order to use the Winsorize FUN with set min/max vals, Winsorize must be capitalized :)
# original: easeimm$siga_frc
# cleaned:
easeimm$siga_w <- Winsorize(easeimm$siga_frc, minval = siga_lo_limit, maxval = siga_up_limit, na.rm = TRUE)

easeimm$il1b_w <- Winsorize(easeimm$ln_il1b, minval = il1b_lo_limit, maxval = il1b_up_limit, na.rm = TRUE)

easeimm$crp_w <- Winsorize(easeimm$crp, minval = crp_lo_limit, maxval = crp_up_limit, na.rm = TRUE)

# Code from here to end of chunk thank you to Johnny
# Put both the original data and the cleaned data in the same df:
df_siga <- data.frame(type=c(rep.int("easeimm$siga_frc", length(easeimm$siga_frc)),
                        rep.int("easeimm$siga_w",  length(easeimm$siga_w))),
                data=c(easeimm$siga_frc, easeimm$siga_w))

df_il1b <- data.frame(type=c(rep.int("easeimm$ln_il1b", length(easeimm$il1b)),
                        rep.int("easeimm$il1b_w",  length(easeimm$il1b_w))),
                data=c(easeimm$il1b, easeimm$il1b_w))

df_crp <- data.frame(type=c(rep.int("easeimm$crp", length(easeimm$crp)),
                        rep.int("easeimm$crp_w",  length(easeimm$crp_w))),
                data=c(easeimm$crp, easeimm$crp_w))

# This is to check it out but I'd probably change it to two boxplots:
ggplot(df_siga, aes(x=data, y=as.numeric(type), color=type))+
  geom_point(position=position_jitter(height=0.5))

ggplot(df_il1b, aes(x=data, y=as.numeric(type), color=type))+
  geom_point(position=position_jitter(height=0.5))

ggplot(df_crp, aes(x=data, y=as.numeric(type), color=type))+
  geom_point(position=position_jitter(height=0.5))

# identify the points that were changed to the min and max
# (aka those points that are equal to the min and max but weren't before)

cleaned_siga <- na.omit(easeimm$siga_w)
original_siga <- na.omit(easeimm$siga_frc)

min_pts_siga <- which(cleaned_siga == min(cleaned_siga) & original_siga != min(cleaned_siga))
max_pts_siga <- which(cleaned_siga == max(cleaned_siga) & original_siga != max(cleaned_siga))

cleaned_il1b <- na.omit(easeimm$il1b_w)
original_il1b <- na.omit(easeimm$il1b)

min_pts_il1b <- which(cleaned_il1b == min(cleaned_il1b) & original_il1b != min(cleaned_il1b))
max_pts_il1b <- which(cleaned_il1b == max(cleaned_il1b) & original_il1b != max(cleaned_il1b))

cleaned_crp <- na.omit(easeimm$crp_w)
original_crp <- na.omit(easeimm$crp)

min_pts_crp <- which(cleaned_crp == min(cleaned_crp) & original_crp != min(cleaned_crp))
max_pts_crp <- which(cleaned_crp == max(cleaned_crp) & original_crp != max(cleaned_crp))

# NOTE: I haven't tested anything below because I don't have a variable with more than one outlier...

# then rank them
# rank will give you smallest -> largest ranks, 
# so to get least-> most outlierish we just reverse min_ranks
min_ranks_siga <- rank(original_siga[min_pts_siga])
min_ranks_siga <- max(min_ranks_siga) - min_ranks_siga + 1
max_ranks_siga <- rank(original_siga[max_pts_siga])

min_ranks_il1b <- rank(original_il1b[min_pts_il1b])
min_ranks_il1b <- max(min_ranks_il1b) - min_ranks_il1b + 1
max_ranks_il1b <- rank(original_il1b[max_pts_il1b])

min_ranks_crp <- rank(original_crp[min_pts_crp])
min_ranks_crp <- max(min_ranks_crp) - min_ranks_crp + 1
max_ranks_crp <- rank(original_crp[max_pts_crp])

# now you can replace them with whatever you want

rank_preserving_siga <- cleaned_siga
rank_preserving_siga[min_pts_siga] <- rank_preserving_siga[min_pts_siga]-(increment * min_ranks_siga)
rank_preserving_siga[max_pts_siga] <- rank_preserving_siga[max_pts_siga]+(increment * max_ranks_siga)

rank_preserving_il1b <- cleaned_il1b
rank_preserving_il1b[min_pts_il1b] <- rank_preserving_il1b[min_pts_il1b]-(increment * min_ranks_il1b)
rank_preserving_il1b[max_pts_il1b] <- rank_preserving_il1b[max_pts_il1b]+(increment * max_ranks_il1b)

rank_preserving_crp <- cleaned_crp
rank_preserving_crp[min_pts_crp] <- rank_preserving_crp[min_pts_crp]-(increment * min_ranks_crp)
rank_preserving_crp[max_pts_crp] <- rank_preserving_crp[max_pts_crp]+(increment * max_ranks_crp)

# check out what we did
outliers_siga <- c(min_pts_siga, max_pts_siga)
comparison_siga <- cbind(original_siga[outliers_siga],
                    cleaned_siga[outliers_siga],
                    rank_preserving_siga[outliers_siga])

outliers_il1b <- c(min_pts_il1b, max_pts_il1b)
comparison_il1b <- cbind(original_il1b[outliers_il1b],
                    cleaned_il1b[outliers_il1b],
                    rank_preserving_il1b[outliers_il1b])

outliers_crp <- c(min_pts_crp, max_pts_crp)
comparison_crp <- cbind(original_crp[outliers_crp],
                    cleaned_crp[outliers_crp],
                    rank_preserving_crp[outliers_crp])

# reorder by the first column
comparison_siga <- comparison_siga[order(comparison_siga[,1]),]

comparison_il1b <- comparison_il1b[order(comparison_il1b[,1]),]

comparison_crp <- comparison_crp[order(comparison_crp[,1]),]

```




# Step 8: Report how many samples were outliers (may include previously OOR values)
```{r}


# Count outliers and report these numbers
siga_num_right_outliers <- length(na.omit(easeimm$siga_frc [easeimm$siga_frc > siga_up_limit])) 
siga_num_left_outliers <- length(na.omit(easeimm$siga_frc [easeimm$siga_frc < siga_lo_limit])) 


# Visualize again after fixing outliers & double check skew, etc
hist(easeimm$siga_w)
hist(easeimm$il1b_w)
hist(easeimm$crp_w)

ggplot(easeimm, aes(x = "", y = siga_w)) +   
  geom_boxplot() +
  ylab("SIgA (pg/ml)") +
  ggtitle("EASE SIgA - winsorized") +
  geom_smooth(method='lm', color="black")

ggplot(easeimm, aes(x = "", y = il1b_w)) +   
  geom_boxplot() +
  ylab("IL-1B (pg/ml)") +
  ggtitle("EASE IL-1B - winsorized") +
  geom_smooth(method='lm', color="black")

ggplot(easeimm, aes(x = "", y = crp_w)) +   
  geom_boxplot() +
  ylab("CRP (pg/ml)") +
  ggtitle("EASE CRP - winsorized") +
  geom_smooth(method='lm', color="black")

siga_w_sk <- skewness(easeimm$siga_w, na.rm = TRUE)
siga_w_ku <- kurtosis(easeimm$siga_w, na.rm = TRUE)
il1b_w_sk <- skewness(easeimm$il1b_w, na.rm = TRUE)
il1b_w_ku <- kurtosis(easeimm$il1b_w, na.rm = TRUE)
crp_w_sk <- skewness(easeimm$crp_w, na.rm = TRUE)
crp_w_ku <- kurtosis(easeimm$crp_w, na.rm = TRUE)


# If your skew/kurtosis are fine, no need to use non-parametric tests now. And if the immune variables are dependent instead of independent variables, you may not need to worry at all.

```

#Step 9: Correlations between markers
```{r}
#If you are analyzing more than one immune marker, ESPECIALLY if you are utilizing multiplex methods, do examine and report multicollinearity. This will inform you on the best final preprocessing methods regarding possible PCA/latent variables or multiple regression tests. However, be careful about combining these markers into latent variables if the markers each have a specific and distinct function (which they often do). If you have a specific hypothesis about the function/association of each immune marker, then analyze each of them. However, you may want to consider correcting for multiple comparisons.

# Take the columns that are your final, flow corrected, transformed, and winsorized immune marker variables:
immmarkers_only <- easeimm[c(17:19)]

imm_corr <- rcorr(as.matrix(immmarkers_only))
#Report these correlations ($r is the correlation, $n the number of obs, $P the p-values):
imm_corr

#Visualize the correlations:
corrplot(imm_corr$r, type = "upper", order = "hclust", 
         tl.col = "black", tl.srt = 45)
```

# Oh and don't forget to export everything!
```{r}
# I'm writing to a different folder to differentiate between "processed" and "scored" data
write.csv(easeimm, file = "A:/Studies/EASE/output/scored_immunology_data/immunology_cleaned.csv")

# You want to save all the summary stats, too probably

write.csv(imm_corr$r, file = "A:/Studies/EASE/output/summary_immunology_data/immunology_corr_r.csv")
write.csv(imm_corr$n, file = "A:/Studies/EASE/output/summary_immunology_data/immunology_corr_n.csv")
write.csv(imm_corr$P, file = "A:/Studies/EASE/output/summary_immunology_data/immunology_corr_p.csv")

easeimm_summary <- data.frame("siga_sk" = siga_sk, "siga_ku" = siga_ku, "ln_siga_sk" = ln_siga_sk, "ln_siga_ku" = ln_siga_ku, "siga_w_sk" = siga_w_sk, "siga_w_ku" = siga_w_ku, "il1b_sk" = il1b_sk, "il1b_ku" = il1b_ku, "ln_il1b_sk" = ln_il1b_sk, "ln_il1b_ku" = ln_il1b_ku, "il1b_w_sk" = il1b_w_sk, "il1b_w_ku" = il1b_w_ku, "crp_sk" = crp_sk, "crp_ku" = crp_ku, "ln_crp_sk" = ln_crp_sk, "ln_crp_ku" = ln_crp_ku, "crp_w_sk" = crp_w_sk, "crp_w_ku" = crp_w_ku)

easeimm_summary$siga_raw_mean <- mean(easeimm$siga, na.rm = TRUE)
easeimm_summary$siga_raw_sd <- sd(easeimm$siga, na.rm = TRUE)
easeimm_summary$siga_w_mean <- mean(easeimm$siga_w, na.rm = TRUE)
easeimm_summary$siga_w_sd <- sd(easeimm$siga_w, na.rm = TRUE)

easeimm_summary$il1b_raw_mean <- mean(easeimm$il1b, na.rm = TRUE)
easeimm_summary$il1b_raw_sd <- sd(easeimm$il1b, na.rm = TRUE)
easeimm_summary$il1b_w_mean <- mean(easeimm$il1b_w, na.rm = TRUE)
easeimm_summary$il1b_w_sd <- sd(easeimm$il1b_w, na.rm = TRUE)

easeimm_summary$crp_raw_mean <- mean(easeimm$crp, na.rm = TRUE)
easeimm_summary$crp_raw_sd <- sd(easeimm$crp, na.rm = TRUE)
easeimm_summary$crp_w_mean <- mean(easeimm$crp_w, na.rm = TRUE)
easeimm_summary$crp_w_sd <- sd(easeimm$crp_w, na.rm = TRUE)

write.csv(easeimm_summary, file = "A:/Studies/EASE/output/summary_immunology_data/immunology_summary.csv")
```

